---
title: "PHStatsII-HW3b"
author: "Marty Ross"
date: "2022-10-02"
output: pdf_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
setwd("~/SpatialAnalysis/01-2022_PHStatsII/Assignments/A3/A3ptB")
casData <- read.csv("casData.csv")
# head(casData)
# Part A:  GPA by drinking
# class(casData$DRINK)
# table(casData$DRINK)
# make DRINK a factor with labels (helpful for visual displays, cuts down on code 
# needed to label groups)
casData$DRINK <- factor(casData$DRINK, levels = c(0, 1), 
                        labels = c("Non-drinker", "Drinker"))
```
# Exercise 1:
## Part A: What is the unadjusted relationship between GPA and consumption of alcohol?
Q1: Simple Boxplot of GPA by drinking status
```{r}
ggplot(casData, aes(x = DRINK, y = GPA, fill = DRINK)) +
  geom_boxplot() + 
  theme_bw(base_size = 14) +
  labs(title = 'Q1: GPA by Drinking Status') +
  theme(axis.title.x = element_blank(),
        legend.position = 'none')
```

Q2-3. Regress by Drinking Status
```{r}
model1 <- lm(GPA ~ DRINK, data = casData)
# summary(model1)
co1 <- model1$coefficients
cat(paste0('Q2. Estimated mean difference in GPA for students who drink versus those\n',
          'who do not:\n  ', round(co1[2], 2)))
```
```{r}
conf1 <- confint(model1)
cat(paste0('Q3. 95% Confidence interval difference in GPA for students who drink\n',
           'versus those who do not:\n  (', paste(round(conf1[2,], 2), collapse = ', '), ')'))
```
## Part B: What is the relationship between GPA and consumption of alcohol, when adjusted for a student’s age?
Q4: A scatterplot showing the relationship between GPA and a student’s age
```{r}
ggplot(casData, aes(x = age, y = GPA, color = DRINK)) +
  geom_point() +
  theme_bw() +
  labs(x = 'Age', color = 'Drinking\nStatus',
       title = 'Q4: GPA by Age',
       subtitle = 'Harvard College Alchol Study')

```

Q5. Side-by-side boxplots showing the difference in age between students who drink and students who do not drink.  
<em>Note: I wasn't certain I was fulfilling this correctly based on the starter code and how the question was phrased, so I'm giving a couple options.</em>
```{r}
casData <- casData %>% mutate(agef = factor(age))

ggplot(casData, aes(x = agef, y = GPA, fill = agef)) +
  geom_boxplot() +
  facet_wrap(facets = casData$DRINK) +
  theme_bw(base_size = 14) +
  theme(legend.position = 'none') +
  labs(x = 'Age', title = 'GPA by Age and Drinking Status')

ggplot(casData, aes(x = DRINK, y = age, fill = DRINK)) +
  geom_boxplot() +
  theme_bw(base_size = 14) +
  theme(legend.position = 'none',
        axis.title.x = element_blank()) +
  labs(y = 'Age',
       title = 'Age by Drinking Status')

```

Q6-9 Multiple Regression on GPA, Drinking + age
```{r}
model2 <- lm(GPA ~ DRINK + age, data = casData)
# summary(model2)
co2 <- coef(model2)
conf2 <- confint(model2)
```

```{r}
# The estimated mean GPA for 20-year-old students who drink
cat(paste0('Q6. The estimated mean GPA for 20-year-old students who drink:\n  ', 
    round(co2[1] + co2[2]*1 + co2[3]*20,2)))
cat(paste0('Q7. The estimated mean GPA for 20-year-old students who do not drink:\n  ', 
    round(co2[1] + co2[2]*0 + co2[3]*20, 2)))
cat(paste0('Q8. The estimated mean difference in GPA for students who drink, compared\n',
    'to students of the same age who do not:\n  ', round(co2[2], 2)))
cat(paste0('Q9. A 95% confidence interval for this estimated difference:\n  (',
    paste(round(conf2[2,],2), collapse = ', '), ')'))
```

## Part C: What is the relationship between GPA and consumption of alcohol, when adjusted for a student’s year in school?

Q10. Side-by-side boxplots showing the relationship between GPA and a student’s year in school. [Hint: To include school year as a categorical variable, you must use as.factor(schoolyr) instead of just schoolyr in your R code.]
```{r}
#  created factor version of year in school
casData <- casData %>%
  mutate(schoolyrf = factor(schoolyr, levels = c(1,2,3,4,5),
                           labels = c("Freshman", "Sophomore", "Junior", "Senior", 
                                      "5th year plus")))
ggplot(casData, aes(x = schoolyrf, y = GPA, fill = schoolyrf)) +
  geom_boxplot() +
  theme_bw(base_size = 14) +
  theme(legend.position = 'none') +
  labs(x = 'Year in School',
       title = 'GPA by Year in School')

```

Q11. A table showing the number and proportions of students in each school year who drink/don’t drink
```{r}
yrdrnk_df <- casData %>%
  count(schoolyrf, DRINK) %>%
  group_by(schoolyrf) %>%
  spread(DRINK, n) %>%
  rename_with(~ paste0("# ", .x))
yrdrnk_df2 <- casData %>%
  count(schoolyrf, DRINK) %>%
  group_by(schoolyrf) %>%
  mutate(prop = paste0(round(n/sum(n)*100,1),"%")) %>%
  select(-n) %>%
  spread(DRINK, prop) %>%
  rename_with(~ paste0("% ", .x))
yrdrnk_df <- cbind(yrdrnk_df, yrdrnk_df2[2:3])
colnames(yrdrnk_df)[1] <- 'School Year'
knitr::kable(yrdrnk_df, format = 'latex', align = 'c', 
             caption = 'Q11. Proportions of Drinkers by School Year') 
```

Q12-15. Model GPA by drinking status and school year
```{r}
# multiple regression of GPA on drinking and school year
model3 <- lm(GPA ~ DRINK + schoolyrf, data = casData)
# summary(model3)
co3 <- coef(model3)
conf3 <- confint(model3)
```

```{r}
cat(paste0('Q12. The estimated mean GPA for freshman students who drink:\n  ', 
    round(co3[1] + co3[2]*1, 2), '\n'))
cat(paste0('Q13. The estimated mean GPA for senior students who drink:\n  ', 
    round(co3[1] + co3[2]*1 + co3[5]*1, 2), '\n'))
cat(paste0('Q14. The estimated mean difference in GPA for students who drink, compared\n',
    'to students of the same school year who do not:\n  ', round(co3[2],2), '\n'))
cat(paste0('Q15. A 95% confidence interval for this estimated difference:\n  (', 
    paste(round(conf3[2,], 2), collapse = ', '),  ')'))
```

## Part D: What is the relationship between GPA and consumption of alcohol, when adjusted for a student’s sex?
Q16. Side-by-side boxplots showing the difference in GPA between male and female students
```{r}
casData <- casData %>%
  mutate(Sex = factor(male, levels = c(0, 1), labels = c('Female', 'Male')))
ggplot(casData, aes(x = Sex, y = GPA, fill = Sex)) +
  geom_boxplot() +
  theme_bw(base_size = 14) +
  theme(legend.position = 'none',
        axis.title.x = element_blank()) +
  labs(title = 'Q16. GPA by Sex')

```

Q17. A table showing the number and proportions of students of each sex who drink/don’t drink
```{r}
sxdrnk_df <- casData %>%
  count(Sex, DRINK) %>%
  group_by(Sex) %>%
  spread(DRINK, n) %>%
  rename_with(~ paste0("# ", .x))
sxdrnk_df2 <- casData %>%
  count(Sex, DRINK) %>%
  group_by(Sex) %>%
  mutate(prop = paste0(round(n/sum(n)*100,1),"%")) %>%
  select(-n) %>%
  spread(DRINK, prop) %>%
  rename_with(~ paste0("% ", .x))
sxdrnk_df <- cbind(sxdrnk_df, sxdrnk_df2[2:3])
colnames(sxdrnk_df)[1] <- 'Drinking Status'
knitr::kable(sxdrnk_df, 'latex', align = 'c', 
             caption = 'Q17. Proportions of Drinkers by Sex') 

```

```{r}
model4 <- lm(GPA ~ DRINK + male, data=casData)
#summary(model4)
co4 <- coef(model4)
conf4 <- confint(model4)
```


```{r}
cat(paste0('Q18. The estimated mean difference in GPA for students who drink, compared to\n',
    'students of the same sex who do not:\n  ', round(co4[2], 2), '\n'))
cat(paste0('Q19. A 95% confidence interval for this estimated difference:\n  (', 
    paste(round(conf4[2,], 2), collapse = ', '), ')\n'))
cat(paste0('Q20. The estimated mean difference in GPA for male students, compared to female\n',
    'students of the same drinking status:\n  ', round(co4[3],2), '\n'))
cat(paste0('Q21. A 95% confidence interval for this estimated difference:\n  (', 
    paste(round(conf4[3,], 2), collapse = ', '), ')'))
```

Q22. According to the College Alcohol Study from the Harvard School of Public Health, it was found that students who drank, defined as having drank alcohol in the past 30 days, were observed as having a 0.19 lower mean GPA (95% CI: -0.29,-0.09) than students who did not drink alcohol when controlling for year in school, a significant finding at the 0.05 confidence level. 

Q23. See above

# Exercise 2
## Part A: What is the unadjusted relationship between high expenditures and smoking?
```{r}
nmesData <- read.csv("nmesData.csv")
# head(nmesData)
nmesData$eversmk <- factor(nmesData$eversmk, levels = c(0, 1), 
                           labels = c('Have not smoked','Have smoked'))
nmesData$highexp <- factor(nmesData$highexp, levels = c(0, 1), 
                           labels = c('Not high expenditures','High expenditures'))
# table of high expenditures indicator by smoking status
smkhiexp_df <- nmesData %>% 
  count(eversmk, highexp) %>%
  group_by(eversmk) %>%
  spread(eversmk, n)
smkhiexp_df2 <- nmesData %>% 
  count(eversmk, highexp) %>%
  group_by(eversmk) %>%
  mutate(prop = paste0(round(n / sum(n)*100,1),"%")) %>%
  select(-n) %>%
  spread(eversmk, prop)
smkhiexp_df <- cbind(smkhiexp_df, smkhiexp_df2[,2:3])
colnames(smkhiexp_df)[1] <- 'Health Care Costs'
knitr::kable(smkhiexp_df, format = 'latex', align = 'c', 
             caption = 'Q24. Counts, Proportions of HiExp by Smoking status') 
```

```{r}
cat(paste0('Q24a. Number and proportion of smokers with high health expenditures\n  ',
           smkhiexp_df[2,3], ', ', smkhiexp_df[2,5]), '\n')
cat(paste0('Q24b. Number and proportion of nonsmokers with high health expenditures\n  ',
           smkhiexp_df[2,2], ', ', smkhiexp_df[2,4]))
```

```{r}
model5 <- glm(highexp ~ eversmk, family=binomial(link=logit), data=nmesData)
#summary(model5)

co5 <- coefficients(model5)
conf5 <- confint(model5)
```

```{r}
od_smk <- exp(co5[1] + co5[2]*1)
cat(paste0('Q25. The estimated probability of a high expenditure for a smoker\n  ',
           round(od_smk / (1 + od_smk), 3)), '\n')
cat(paste0('Q26. The estimated probability of a high expenditure for a non-smoker\n  ',
           round(exp(co5[1] / (1 + exp(co5[1]))), 3)), '\n')
cat(paste0('Q27. The estimated odds ratio of a high expenditure, comparing smokers to ',
           'nonsmokers\n  ', round(exp(co5[2]), 3)), '\n')
cat(paste0('Q28. A 95% confidence interval for this true odds ratio of high expenditure, ',
           '\ncomparing smokers to nonsmokers\n  (',
           paste(round(exp(conf5[2,]),3), collapse = ', '), ')'))
```

## Part B: What is the relationship between high expenditures and smoking when adjusted for an individual’s age?
```{r}
model6 <- glm(highexp ~ eversmk + age, family=binomial(link=logit), data=nmesData)
# summary(model6)
co6 <- coefficients(model6)
conf6 <- confint(model6)
od_smk50 <- exp(co6[1] + co6[2]*1 + co6[3]*50)
cat(paste0('Q29. The estimated probability of a high expenditure for a 50-year-old smoker\n  ',
           round(od_smk50 / (1 + od_smk50), 3), '\n'))
od_nosmk60 <- exp(co6[1] + co6[3]*60)
cat(paste0('Q30. The estimated probability of a high expenditure for a 60-year-old ', 
           'non-smoker\n  ', round(od_nosmk60 / (1 + od_nosmk60), 3), '\n'))
cat(paste0('Q31. The estimated odds ratio of a high expenditure, comparing smokers to ',
           'nonsmokers\nof the same age\n  ', round(exp(co6[2]),3), '\n'))
cat(paste0('Q32. A 95% confidence interval for this estimated odds ratio\n  (',
           paste(round(exp(conf6[2, ]),3), collapse = ', '),')\n'))
cat(paste0('Q33. The estimated odds ratio of a high expenditure for two groups of the ',
           'same\nsmoking status but whose age differs by 10 years\n  ',
           round(exp(co6[3]*10),3), '\n'))
cat(paste0('Q34. A 95% confidence interval for this estimated odds ratio\n  (',
           paste(round(exp(conf6[3,]*10), 3), collapse = ', '),')'))
```
```{r}
paste0(round(exp(co6[2]),3), ' (', paste(round(exp(conf6[2,]),3), collapse = ', '), ')')
```

Q35. According to data collected from 4078 individuals in the 1987 National Medical Expenditures Survey (NMES), the status of ever having smoked is associated with an estimated 54% greater odds of higher health care expenditures (95% CI: 14.9%, 107.3%) when adjusted for age. This is a significant finding at the 0.05 level.

Q36. See above

